A density ratio framework for evaluating the utility of synthetic data

### 2025-09-28 paper/paper.tex
Update for new submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\newcommand{\Xobs}{\mathbf{X}_{\text{obs}}}
\newcommand{\Xsyn}{\mathbf{X}_{\text{syn}}}
\newcommand{\dist}{\mathcal{P}}
\newcommand{\pobs}{p_{\text{obs}}}
\newcommand{\psyn}{p_{\text{syn}}}
\newcommand{\nobs}{n_{\text{obs}}}
\newcommand{\nsyn}{n_{\text{syn}}}
\newcommand{\bx}{\mathbf{x}}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{float}
\floatplacement{table}{t}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A density ratio framework for evaluating the utility of synthetic data},
  pdfauthor={Thom Benjamin Volker; Peter-Paul de Wolf; Erik-Jan van Kesteren},
  pdfkeywords={synthetic data, utility, density ratio, privacy,
disclosure limitation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

Openly accessible research data accelerates scientific progress
tremendously. Open data allows third-party researchers to answer
research questions with already collected data, freeing up resources
that would otherwise be devoted to data collection
(\citeproc{ref-ramachandran_open_2021}{Ramachandran, Bugbee, and Murphy
2021}). Sharing data in combination with code allows others to validate
research findings and build upon the work
(\citeproc{ref-obels_analysis_2020}{Obels et al. 2020};
\citeproc{ref-crosas_automating_2015}{Crosas et al. 2015}). Students can
benefit from open data, as it fosters education with realistic data
(\citeproc{ref-atenas_open_2015}{Atenas, Havemann, and Priego 2015}), as
well as the general public, through stimulating citizen science projects
(\citeproc{ref-newman_future_2012}{Newman et al. 2012}). However, making
data openly available is often (rightfully) hampered by official
legislation, like the General Data Protection Regulation (GDPR;
\citeproc{ref-gdpr}{European Parliament and Council of the European
Union 2016}), and general privacy concerns. In the worst case, sharing
data may cause harm to individuals or organizations, which may withhold
these entities from participating in future research. These privacy
constraints have been named among the biggest hurdles in the advancement
of computational social science (\citeproc{ref-lazer_css_2009}{Lazer et
al. 2009}), and among top reasons for companies to not share their data
with researchers (\citeproc{ref-fpf_2017}{Future of Privacy Forum
2017}).

@@

To overcome the issues related to traditional global utility measures,
we propose to use the density ratio estimation framework
(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}) as a way of evaluating utility. Intuitively, if two
data sets have similar multivariate distributions, the density ratio
should be close to one over the range of the data. If the distributions
of the observed and synthetic data are very different, the density ratio
should be far from one at those regions where the distributions differ.
As density estimation is known to be a difficult problem, the density
ratio estimation framework provides techniques to directly estimate the
density ratio, rather than the two separate densities, in a
non-parametric way
(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}). These non-parametric estimation techniques come with
automatic model specification, which mitigates the issue of model
specification. This functionality is implemented in the
\texttt{R}-package \texttt{densityratio}
(\citeproc{ref-densityratio}{Volker 2023}). Importantly, the density
ratio is estimated over the entire range of the data, which provides
measures of utility at every (possible) point in the data space. This
point-specific quantification of utility turns out to be a useful
side-product, as it allows to reweigh analyses on synthetic data when
further improving the utility directly is not possible.

In the remainder of the article, we introduce the density ratio
framework and the associated estimation techniques, and connect the
framework to traditional utility measures as the \(pMSE\) and the
Kullback-Leibler divergence. We then present simulations to demonstrate
the performance of density ratio estimation in stylized settings and
compare it to traditional utility measures. Subsequently, we apply
density ratio estimation in a case study where we evaluate the utility
of multiple synthetic versions of the U.S. Current Population Survey. We
conclude with a discussion of the results, highlight the strengths and
weaknesses of the density ratio framework, and provide recommendations
for future research.

@@

Global utility measures typically attempt to quantify the distributional
similarity between the observed and synthetic data samples. The
intuition is that two data sets are realizations from the same
underlying distribution, the data sets can be used for the same
purposes, and analyses on the two data sets should yield similar
results. One way to evaluate distributional similarity is by assessing
whether a classification model can tell samples from the two
distributions apart (see \citeproc{ref-kim_classification_2021}{Kim et
al. 2021}, who formalize the connection between classification accuracy
and two-sample testing). Hence, if a classification model can
distinguish between the observed and synthetic data with high accuracy,
the distributional similarity is low, and so is the global utility. If a
classifier cannot distinguish between the observed and synthetic data,
one would conclude that the global utility is high. The propensity score
mean-squared error (\(pMSE\)), introduced by Woo et al.
(\citeproc{ref-Woo_global_2009}{2009}) and further developed in Snoke et
al. (\citeproc{ref-snoke_utility_2018}{2018}), formalizes this
intuition. Let \(I_i\) denote an indicator variable that equals \(1\) if
observation \(i\) (\(i \in 1, \dots, N\), \(N = \nobs + \nsyn\)) belongs
to the synthetic data, and \(0\) otherwise. We then train a classifier
that outputs the predicted probability of observation \(i\) being a
synthetic record \(\hat{\pi}_i\) based on the observation's scores on
the variables (this can be the set of all variables, but also a subset).
From these, we can calculate the utility statistic
\begin{equation}\phantomsection\label{eq-pMSE}{
pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{\pi}_i - \frac{\nsyn}{N}\Big)^2,
}\end{equation} which ought to be smaller when the synthetic data is
more like the observed data. Crucially, the \(pMSE\) depends on the
classification model used and increases in the flexibility of the
classification model, making it prone to overfitting and hard to
interpret. To combat these issues, Snoke et al.
(\citeproc{ref-snoke_utility_2018}{2018}) suggests to compare the
\(pMSE\)-value with its expected value under the null hypothesis that
the observed and synthetic data are indistinguishable. Provided that the
classification model is a logistic regression model with \(k\)
parameters (including the intercept), Snoke et al.
(\citeproc{ref-snoke_utility_2018}{2018}) show that the expected
\(pMSE\) is given by \[
\mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
\Big(\frac{\nsyn}{N}\Big).
\] For other classification models, the expectation can be approximated
through a resampling procedure. Accordingly, the \(pMSE\)-ratio is given
by \[
pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]}.
\] Apart from the \(pMSE\), several other measures can be constructed
from the estimated propensity scores, such as the percentage of records
correctly predicted (\citeproc{ref-raab2021assessing}{Gillian M. Raab,
Nowok, and Dibben 2021}) or the Kolmogorov-Smirnov statistic
(\citeproc{ref-Bowen_differentially_2021}{Bowen, Liu, and Su 2021}),
both which are strongly correlated with the \(pMSE\)
(\citeproc{ref-raab2021assessing}{Gillian M. Raab, Nowok, and Dibben
2021}).

@@

The density ratio estimation framework was originally developed in the
machine learning community for the comparison of two probability
distributions (for an overview, see
\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}). The framework has been shown to be applicable to
prediction (\citeproc{ref-sugiyama_conditional_2010}{Sugiyama et al.
2010}; \citeproc{ref-sugiyama_classification_2010}{Sugiyama 2010}),
outlier detection (\citeproc{ref-shohei_dre_outlier_2008}{Hido et al.
2008}), change-point detection in time-series
(\citeproc{ref-liu_change_2013}{Liu et al. 2013}), importance weighting
under domain adaptation (i.e., sample selection bias;
\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009}),
and two-sample homogeneity tests
(\citeproc{ref-sugiyama_lstst_2011}{Sugiyama, Suzuki, et al. 2011}). The
general idea of density ratio estimation is depicted in
Figure~\ref{fig-dr-plot}, and boils down to comparing two distributions
by modelling the density ratio \(r(\bx)\). In our case, we define
\(r(\bx)\) as the ratio between the probability distributions of the
numerator samples, taken from the observed data distribution,
\(\pobs(\bx)\), and the denominator samples, taken from the synthetic
data distribution, \(\psyn(\bx)\), such that
\begin{equation}\phantomsection\label{eq-dr}{
r(\bx) = \frac{\pobs(\bx)}{\psyn(\bx)}.
}\end{equation}

This specification has the intuitive interpretation that in locations
where the density ratio takes values larger than 1, too few synthetic
observations are generated in that region. Conversely, in locations
where the density ratio takes values smaller than one, too many
synthetic observations are generated in that region (see
Figure~\ref{fig-dr-plot}). We put \(\pobs(\bx)\) in the numerator such
that it plays the role of the actual distribution, while \(\psyn(\bx)\)
serves as the approximating distribution (we briefly come back to this
in the next section). A straightforward approach to estimating
\(r(\bx)\) from samples of \(\pobs(\bx)\) and \(\psyn(\bx)\) would be to
estimate the observed and synthetic data density separately, for example
using kernel density estimation (e.g., see
\citeproc{ref-Scott1992}{Scott 1992} for an overview), and subsequently
compute the ratio of these estimated densities. However, density
estimation is one of the hardest tasks in statistical learning,
unavoidably leading to estimation errors for both densities, especially
in high dimensions
(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}). When subsequently taking the ratio of the estimated
densities, estimation errors tend to be magnified. Direct density ratio
estimation avoids this issue by specifying and estimating a model
directly for the ratio without first estimating the separate densities.
Extensive simulations on a wide variety of tasks showed that this
approach typically outperforms density ratio estimation through kernel
density estimation for each distribution separately, especially when the
dimensionality of the data increases (e.g.,
\citeproc{ref-Kanamori2012}{Kanamori, Suzuki, and Sugiyama 2012b};
\citeproc{ref-shohei_dre_outlier_2008}{Hido et al. 2008};
\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009}).

@@

\includegraphics{paper_files/figure-pdf/fig-AppA-1.pdf}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\newcommand{\Xobs}{\mathbf{X}_{\text{obs}}}
\newcommand{\Xsyn}{\mathbf{X}_{\text{syn}}}
\newcommand{\dist}{\mathcal{P}}
\newcommand{\pobs}{p_{\text{obs}}}
\newcommand{\psyn}{p_{\text{syn}}}
\newcommand{\nobs}{n_{\text{obs}}}
\newcommand{\nsyn}{n_{\text{syn}}}
\newcommand{\bx}{\mathbf{x}}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{float}
\floatplacement{table}{t}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A density ratio framework for evaluating the utility of synthetic data},
  pdfauthor={Thom Benjamin Volker; Peter-Paul de Wolf; Erik-Jan van Kesteren},
  pdfkeywords={synthetic data, utility, density ratio, privacy,
disclosure limitation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

Openly accessible research data accelerates scientific progress
tremendously. Open data allows third-party researchers to answer
research questions with already collected data, freeing up resources
that would otherwise be devoted to data collection
(\citeproc{ref-ramachandran_open_2021}{Ramachandran, Bugbee, and Murphy
2021}). Sharing data with the accompanying with code allows others to
validate research findings and build upon the work
(\citeproc{ref-obels_analysis_2020}{Obels et al. 2020};
\citeproc{ref-crosas_automating_2015}{Crosas et al. 2015}). Students can
benefit from open data, as it fosters education with realistic data
(\citeproc{ref-atenas_open_2015}{Atenas, Havemann, and Priego 2015}),
and so can the general public, through stimulating citizen science
projects (\citeproc{ref-newman_future_2012}{Newman et al. 2012}).
However, making data openly available is (rightfully) hampered by
official legislation, like the General Data Protection Regulation (GDPR;
\citeproc{ref-gdpr}{European Parliament and Council of the European
Union 2016}), and general privacy concerns. In the worst case, sharing
data may cause harm to individuals or organizations, which may withhold
these entities from participating in future research. These privacy
constraints have been named among the biggest hurdles in the advancement
of computational social science (\citeproc{ref-lazer_css_2009}{Lazer et
al. 2009}), and among top reasons for companies to not share their data
with researchers (\citeproc{ref-fpf_2017}{Future of Privacy Forum
2017}).

@@

To overcome the issues related to traditional global utility measures,
we propose to use the density ratio estimation framework
(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}) as a way of evaluating utility. Intuitively, if two
data sets have similar multivariate distributions, the density ratio
should be close to one over the range of the data. If the distributions
of the observed and synthetic data are very different, the density ratio
should be far from one at those regions where the distributions differ.
The density ratio estimation framework approaches the problem by
estimating the density ratio directly in a non-parametric way
(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}). This approach alleviates the challenging problem of
estimating two densities, which may magnify estimation errors when
subsequently taking their ratio. The density ratio framework is made
readily available in the \texttt{R}-package \texttt{densityratio}
(\citeproc{ref-densityratio}{Volker 2023}), which attempts to further
simplify the procedure by implementing automatic hyperparameter
optimization, which mitigates the issue of model specification.
Importantly, the density ratio is estimated over the entire range of the
data, which provides measures of utility at every (possible) point in
the data space. This point-specific quantification of utility turns out
to be a useful side-product, as it allows to reweigh analyses on
synthetic data when further improving the utility directly is not
possible.

In the remainder of the article, we introduce the density ratio
framework and connect it to traditional utility measures as the \(pMSE\)
and the Kullback-Leibler divergence. We then present simulations to
demonstrate the performance of density ratio estimation in stylized
settings and compare it to traditional utility measures. Subsequently,
we apply density ratio estimation in a case study where we evaluate the
utility of multiple synthetic versions of the U.S. Current Population
Survey. We conclude with a discussion of the results, highlight the
strengths and weaknesses of the density ratio framework, and provide
recommendations for future research.

@@

Global utility measures typically attempt to quantify the distributional
similarity between the observed and synthetic data samples. The
intuition is that two data sets are realizations from the same
underlying distribution, the data sets can be used for the same
purposes, and analyses on the two data sets should yield similar
results. One way to evaluate distributional similarity is by assessing
whether a classification model can tell samples from the two
distributions apart (see \citeproc{ref-kim_classification_2021}{Kim et
al. 2021}, who formalize the connection between classification accuracy
and two-sample testing). Hence, if a classification model can
distinguish between the observed and synthetic data with high accuracy,
the distributional similarity is low, and so is the global utility. If a
classifier cannot distinguish between the observed and synthetic data,
one would conclude that the global utility is high. The propensity score
mean-squared error (\(pMSE\)), introduced by Woo et al.
(\citeproc{ref-Woo_global_2009}{2009}) and further developed in Snoke et
al. (\citeproc{ref-snoke_utility_2018}{2018}), formalizes this
intuition. Let \(I_i\) denote an indicator variable that equals \(1\) if
observation \(i\) (\(i \in 1, \dots, N\); \(N = \nobs + \nsyn\)) belongs
to the synthetic data, and \(0\) otherwise. These indicators are used in
a classifier that outputs the predicted probability of observation \(i\)
being a synthetic record \(\hat{\pi}_i\) based on the observation's
scores on the variables (this can be the set of all variables, but also
a subset). From these, the utility statistic is calculated as
\begin{equation}\phantomsection\label{eq-pMSE}{
pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{\pi}_i - \frac{\nsyn}{N}\Big)^2,
}\end{equation} which ought to be smaller when the synthetic data is
more like the observed data. Crucially, the \(pMSE\) depends on the
classification model used and increases in the flexibility of the
classification model, making it prone to overfitting and hard to
interpret. To combat these issues, Snoke et al.
(\citeproc{ref-snoke_utility_2018}{2018}) suggests to compare the
\(pMSE\)-value with its expected value under the null hypothesis that
the observed and synthetic data are indistinguishable. Provided that the
classification model is a logistic regression model with \(k\)
parameters (including the intercept), Snoke et al.
(\citeproc{ref-snoke_utility_2018}{2018}) show that the expected
\(pMSE\) is given by \[
\mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
\Big(\frac{\nsyn}{N}\Big).
\] For other classification models, the expectation can be approximated
through a resampling procedure. Accordingly, the \(pMSE\)-ratio is given
by \[
pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]}.
\] Apart from the \(pMSE\), several other measures can be constructed
from the estimated propensity scores, such as the percentage of records
correctly predicted (\citeproc{ref-raab2021assessing}{Gillian M. Raab,
Nowok, and Dibben 2021}) or the Kolmogorov-Smirnov statistic
(\citeproc{ref-Bowen_differentially_2021}{Bowen, Liu, and Su 2021}),
both which are strongly correlated with the \(pMSE\)
(\citeproc{ref-raab2021assessing}{Gillian M. Raab, Nowok, and Dibben
2021}).

@@

The density ratio estimation framework was originally developed in the
machine learning community for the comparison of two probability
distributions (for an overview, see
\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}). The framework has been shown to be applicable to
prediction (\citeproc{ref-sugiyama_conditional_2010}{Sugiyama et al.
2010}; \citeproc{ref-sugiyama_classification_2010}{Sugiyama 2010}),
outlier detection (\citeproc{ref-shohei_dre_outlier_2008}{Hido et al.
2008}), change-point detection in time-series
(\citeproc{ref-liu_change_2013}{Liu et al. 2013}), importance weighting
under domain adaptation (i.e., sample selection bias;
\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009}),
and two-sample homogeneity tests
(\citeproc{ref-sugiyama_lstst_2011}{Sugiyama, Suzuki, et al. 2011}). The
general idea of density ratio estimation is depicted in
Figure~\ref{fig-dr-plot}, and boils down to comparing two distributions
by modelling the density ratio \(r(\bx)\). In our case, we define
\(r(\bx)\) as the ratio between the probability distributions of the
numerator samples, taken from the observed data distribution,
\(\pobs(\bx)\), and the denominator samples, taken from the synthetic
data distribution, \(\psyn(\bx)\), assumed to have common support, such
that \begin{equation}\phantomsection\label{eq-dr}{
r(\bx) = \frac{\pobs(\bx)}{\psyn(\bx)}.
}\end{equation} This specification has the intuitive interpretation that
in locations where the density ratio takes values larger than 1, too few
synthetic observations are generated in that region. Conversely, in
locations where the density ratio takes values smaller than one, too
many synthetic observations are generated in that region (see
Figure~\ref{fig-dr-plot}). We put \(\pobs(\bx)\) in the numerator such
that it plays the role of the actual distribution, while \(\psyn(\bx)\)
serves as the approximating distribution (we briefly come back to this
in the next section). A straightforward approach to estimating
\(r(\bx)\) from samples of \(\pobs(\bx)\) and \(\psyn(\bx)\) would be to
estimate the observed and synthetic data density separately, for example
using kernel density estimation (e.g., see
\citeproc{ref-Scott1992}{Scott 1992} for an overview), and subsequently
compute the ratio of these estimated densities. However, density
estimation is one of the hardest tasks in statistical learning,
unavoidably leading to estimation errors for both densities, especially
in high dimensions
(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
Kanamori 2012a}). When subsequently taking the ratio of the estimated
densities, estimation errors tend to be magnified. Direct density ratio
estimation avoids this issue by specifying and estimating a model
directly for the ratio without first estimating the separate densities.
Extensive simulations on a wide variety of tasks showed that this
approach typically outperforms density ratio estimation through kernel
density estimation for each distribution separately, especially when the
dimensionality of the data increases (e.g.,
\citeproc{ref-Kanamori2012}{Kanamori, Suzuki, and Sugiyama 2012b};
\citeproc{ref-shohei_dre_outlier_2008}{Hido et al. 2008};
\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009}).

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-AppA-1.pdf}}
